---
title: "Notes - R programming"
output: html_notebook
---
# 1 - Week 1 (Background, Getting Started, and Nuts & Bolts)
## 1.1 - Background
### Getting help
Not much - Just google shit!

### R console input and evaluation
Not much

### Data types
R has five basic classes of objects
1. Character
2. Numeric (real numbers)/float
3. integer
4. Complex
5. logical/boolen

Objects include
* Vector - All stuff in a vector needs to be same class - BUT, list can have muliple types of classes. Can be a list that has alot of stuff. Think data-frames.

**Numbers**
Generally R always store stuff as numerics/floats. Entering "numberL" gives a integer.
NaN - not a number or missing value.


**Attributes**
R objects can have attribute
* Names
* Dimensions
* Class
* Lenght
* Other used-defined stuff

Attributes can be seen using the attributes() function.

**Vectors**
Can be made using the c function. I.e.
```{r}
c(1,2,3)
c(TRUE,FALSE)
```

If you take a vector and try to mix different types of objects? R will take the least common object and covert that to a other object.

You can also convert stuff of one object to another like this

```{r}
x <- 0:6 
as.numerix(x)
as.logical(x)
as.character(x)
```

In some cases you might end up with NA's while doing this.

**Lists**
Can contain several types of objects.


```{r}
x <- list(1, "a", TRUE, 1+4i)

print(x)
```

**Matrices**
Is a special type of vector. It is vectors that has several dimensions. It is number of rows and columns.

I can make it like this.
```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
print(m)
```

You can also at a later time make dimensions
```{r}
m <- 1:10
dim(m) <- c(2,5)
print(m)
```
A last way to create a matrix using two allready existing vectors.

```{r}
x <- 1:3
y <- 10:12
cbind(x,y)
```

**Factors**
Factors are used to reperesent chatogorical data. Can either be ordered or unordered.

It is an iteger vercor where each vector has a value. can be created like this:

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"),
            levels = c("yes", "no"))
x
table(x)
unclass(x)
```
 
You can also set the levels of the factors - can be important in linear modeling - See above.

**Missing values**
Missing values are denoted by NA or NaN.

There are several functions one can use in relation to missing values
* is.na()
* is.nan()

NA values have a class - there are integer, character NaN.
 There is some differences (mathematically) betwen the two types of missing values.

**Data frames**
Used to store tabular data. They make up a lot of data types used in statistics.

Data frames are a special type of list. 

Dataframes have alot of different attributes called row.names (subject ID, etc.)

Data frames can usually be created by calling read.table() or read.csv()

can be converted into a data.matrix() = same objects all over

You can manually create a data frame like this

```{r}
x <- data.frame(foo = 1:4, bar = c(T, T, F, F))
```

**Names**
All R-objects can have names. Usable for using readable code. Can be used like this.

```{r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "test")
names(x)
```

Lists and matrices can also have names. When matrecies have names they are called dimnames.

### Reading tabular data
There a many reading functions in r.

* read.table or read.csv - For reading tabular data
* readLines - For reading lines of text file
* Source - For reading R code files
* dget - Reading R code files again
* load - reading in saved workspaces
* unserialize - reading R stuff in binary

Also writing data
* write.table
* writelines
* etc.

read.table is the most used functions. It will usually return a data.frame inside R. There are many important arguments for this command. Some include:
* file - name of file
* header - Header line
* skip - you can skip some rows in the dataset

Dependent on size of dataset you can usually use read.table without need for any arguments.

read.csv is identical to read.table, howver, the csv is used for comma seperated documents.

**Reading large tables**
There are many considerations to make when working with large datasets. You should:

* know the documentation of the function
* make rough calculation of RAM usage.
* Set comment.char = "" - If there is no comments in the file.
* Use colClasses - if not R uses alot of time to figure this out on it self. 

**Textual data**
You can output data from R in text files. Can be good for version control.

Can use functions dumping and dputing. Can be very good, however, takes up alot of space

Can be used like this
```{r}
y <- data.frame(a=1, b= "a")
dput(y)
```

You can read in dput files using dget.

Dump is similar, however, can bs used on several objects in the same time.


### Interface to the outside world
Read different stuff. Examples include:

* file
* gzfile - compressed
* bzfile - compressed
* url - website

You can connect like this

```{r}
con <- file("foo.txt", "r")
data <- read.csv(con)
```

This does the same as read.csv

Other example

```{r}
con <- gzfile("name")
x <- readLines(con, 10)
x
```

### Subsetting
Methods for extrating subsets of R objects

Several types exsists

* [] returns an object of same class as original
* [[]] extract elements of a list or data frame
* $ is used to extract elements of a list or data frame by name.


Example of []
```{r}
x <- c(1:10)
x[1]
x[2]
x[1:4]
x[x > 8]


```

example of [[]]
```{r}
x <- list(foo = 1:4, bar= 0.6)
x [1]

x[[1]]


x ["bar"]


```

See the lectures if you need a specific subsetting function.

There is ways to subset:
* Lists
* Matricies
* Partial matching
* Remove missing values


### Vectorized operations


## 1.2 - Exersices



