---
title: "Lesson"
author: "Anders"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(vroom)
library(here)
library(dplyr)
library(magrittr)
library(fs)
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
devtools::load_all()
```

## 1 - Split-Apply-Combine
This s technique for taking a dataset. Splitting it up for something (i.e. male/female). Doing something to it, and bringing it back. 

* Split = Split data into groups
* Apply = you do something to data based on the split
* Combine = Combine the data back up

```{r, warning = FALSE}
mmash %>% 
  group_by(day) %>% 
  summarise(across(
    c(age, weight, height),
    list(mean = mean, sd = sd, max = max, min = min),
    na.rm = T
  ))
```

## 2 - Pivot longer

```{r}
mmash %>% 
  select(-samples) %>%
  pivot_longer(c(-user_id, -day, -gender))
```

Use pivot_longer() after the group_by() and summarise() we did previously:
```{r}
# Data
mmash %>% 
    group_by(day) %>% 
    summarise(across(
        c(age, weight),
        list(mean = mean, sd = sd), 
        na.rm = TRUE
    ))

# Pivot
mmash %>% 
    group_by(day) %>% 
    summarise(across(
        c(age, weight),
        list(mean = mean, sd = sd), 
        na.rm = TRUE
    )) %>%
  pivot_longer(ends_with(c("mean", "sd")), 
               names_to = c("names", ".value"),
               names_sep = "_")
```
# Pibot wider
We can actually utilize pivot_longer function of naming stuff, to enable us to more easially use pivot_wider.
```{r}
mmash %>% 
    group_by(day) %>% 
    summarise(across(
        c(age, weight),
        list(mean = mean, sd = sd), 
        na.rm = TRUE
    )) %>%
  pivot_longer(ends_with(c("mean", "sd")),
               names_to = c("names", "summary_statistic"),
               names_sep = "_") %>%
  pivot_wider(names_from = summary_statistic,
              values_from = value)
```

Do it using  a funciton
```{r}
tidy_summarize_by_day(mmash, c("age", "weight"))
tidy_summarize_by_day(mmash, c(age, weight))

```
Non-standard evaluation is the thing with {{}}

# You can make a tables in HTML using the kable function in the knitr package
```{r}
tidy_summarize_by_day(mmash, c(age, weight)) %>% 
  knitr::kable(caption = "Nice table")
```

