---
title: "Part 1 - Explore"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
editor_options: 
  markdown: 
    wrap: 72
---

# Load packages

```{r, results='hide', message=FALSE}
# Load tidyvese package package
library(tidyverse)
```

# Chapter 1 - Data visualization using ggplot2

## 1- Introduction

### 1.1 - Notes

No real notes for this section. Book only talks about prerequisites for
running R and excreta.

### 1.2 - Exercises

There is no exercises.

## 2 - First steps

### 2.1 - Notes

We are using the MPG data set. This comes with the tidyverse package.

A standard graphing template is as follows.

``` {.R}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

### 2.2 - Exercises

#### **Question 1 - Run plotting on mpg. What do you see?**

**Answer:** Doing this we see an empty graph. I need to define what I
want to view.

**Output:**

```{r empty ggplot}
ggplot (data = mpg)
```

#### **Question 2 - How many rows and columns are in mpg?**

**Answer:** I can find this using the `str()` (base R), `glimpse()`
(tidyverse), or more specifically `nrow()` and `ncol()`. Or just simply
writing the data in the console.

**Output:**

```{r Rows and columns}
str(mpg)
glimpse(mpg)
nrow(mpg)
ncol(mpg)
mpg
```

It is 234 rows and 11 columns.

#### **Question 3 - what does the dr variable describe?**

**Answer:** I can find out what the `dr` variable describes by looking
trough the documentation for mpg.

**Output:**

```{r look up documentation}
?mpg
```

The variable `drv` describes whether the car is (front wheel drive), r
(rear wheel drive), or 4 (four wheel drive).

#### **Question 4 - make a scatter-plot of hwy vs cyl**

**Answer:** I can do that using `geom_point`, as seen below.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```

#### **Question 5 - What happens if you make a scatter-plot of class vs drv? Why is it not useful?**

**Answer:** See graph below for what happens. Such a graph is not useful
as the type of wheel setup (`drv`) is not dependent on type of car.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = (aes(x = class, y = drv)))
```

## 3 - Aesthetic mappings

### 3.1 - Notes

Aesthetic mapping lets you highlight specific components of the data.
Aesthetics includes things like **size**, **shape** and **color**. One
can use aesthetic mapping like this.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Here is the same example, using size instead.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

This gives us a different aesthetic, however, in this case it is not
advised as it is very difficult to differentiate different groups based
on size. Other things we can map includes: alpha (transparency) and
shape.

Furthermore, we can set aesthetics manually like this.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

Notice that the color is outside of the x and y parentheses, but still
inside the `aes`.

You can look up information on the types of shapes that are available in
R. There is a total of 25.

### 3.2 - Exercises

#### **Question 1 - What's gone wrong with this code? Why are the points not blue?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

**Answer**: The point is not blue because the deceleration for blue
points are inside the `aes` declaration. This creates a new variable in
MPG called blue, which R attempts to color, first color being red.

#### **Question 2 - Which variables in mpg are categorical and continuous?**

**Answer:** One can sometimes figure this out by looking at the
documentation.

```{r}
?mpg
```

In this specific case, it gives us nothing. Therefore, the next step
would be to simply look at the data.

```{r}
mpg
```

While there is no certainty, usually columns coded as characters (`chr`)
can function as categorical while numeric values (`int` and `dbl`) are
continuous.

#### **Question 3 - Map a continuous variable to color, size, and shape and look how they behave compared to categorical values**

**Answer:**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = year, size = cyl))
```

When mapping continuous variables to color, it creates a gradient when
used with continuous variables. Size also work with continuous
variables, however, is not as good and rather scales the variables
instead of having discrete sizes.

Furthermore, a continuous variable cant be mapped to something like
shapes, as there must then be a shape for each unique value.

#### **Question 4 - What happens if you map the same variable to multiple aesthetics?**

**Answer:**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = trans, shape = trans))
```

Both things are mapped, simple as that.

#### **Question 5 - What does the stroke aesthetic do? What shapes does it work with?**

I can see this by looking at the vignette for ggplot2.

```{r}
vignette("ggplot2-specs")
```

From this we learn that stroke can modify the width of the border of
certain shapes (21 - 24).

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = year, size = cyl, stroke = cyl))
```

It can also do this:

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)
```

#### **Question 6 - What happens if you map an aesthetic to something other than a variable name?**

**Answer:**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = displ < 5))
```

Aesthetics can also be mapped to logical expressions.

## 4 - Facets

### 4.1 - Notes

One additional way to show your data is to split your plot into
subplots, also called facets. Facets can be created using the
`facet_wrap()` function. To pass a variable to `facet_wrap()` you need
to cast a variable with a formula (\~).

This can be done line this:

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 1)
```

It is also possible to facet a plot in combination with two variables
using `facet_grid()`. This function creates a x and y axis, sort of like
a heatmap.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ class)
```

It is apparantly possible to make a `facet_grid()` into a
`facet_wrap(nrow = 1)` by writing e.g., `facet_grid(. ~ cyl)`. I do
however not see why would do this instead of a facet wrap. I would argue
that a wrap is more readable.

### 4.2 - Exercises

#### **Question 1 - What happens if you facet on a continuous variable?**

**Answer:**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ cty, nrow = 2)
```

We get so many facets/subplots as there is unique values. So it is not
smart.

#### **Question 2 - What do the empty cells in a plot with facet_grid(drv \~ cyl) mean? How do they relate to this plot?**

**Answer:**

Empty cells means that there are some `drv ~ cyl` combinations that have
no values (hence empty).

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_grid(drv ~ cyl)
```

#### **Question 3 - What plots does the following code make? What does . do?**

**Answer:**

As I wrote in my notes, the . makes essentially a `facet_wrap()` with
only one row or column. The `.` simply states that you do not want the
facets in any dimensions.

Example for subplotting over several columns:

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```

Example for subpotting over several rows.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(. ~ drv)
```

I would argue it is better to use `facet_wrap(ncol = 1)` or
`facet_wrap(nrow = 1)`.

#### **Question 4 - Take this faceted plot. What are the advantages and disadvantages of using faceting instead of color aesthetic?**

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

**Answer:**

[Advantages]{.ul}

-   You split up your view. Make it simpler if you have a larger
    dataset.

-   If you have many variables to show, it can be difficult to see
    differences in shapes and colour.

-   If the cases are somehow independent of each other.

[Disadvantages]{.ul}

-   If you just have a small dataset, it can seem empty.

-   Difficult to compare direct values.

#### **Question 5 - What does `nrow` and `ncol` do in facets? And why does not facet_grid() have a `nrow` and `ncol` variable?**

**Answer:** I will need to look at the documentation of facets to figure
this out.

```{r}
?facet_wrap
```

`ncol()` sets number of columns while `nrow()` sets number of rows. It
is used if you wish to specify your own rows and columns.

These variables are not needed for `facet_grid()` as this function lays
it out in rows and columns dependent on the data/equation selected.

#### **Question 6 - When using `facet_grid()` you should usually put the variable with more unique levels in the columns, why?**

**Answer:** This is probably because it is easier to read if its longer
in the column compared to rows.

## 5 - Geometric Objects

### 5.1 - Notes

You make different plots/graphs in R by using different `geoms`
(geometric object). These can include:

-   geom_point = scatter plot

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

-   geom_smooth = line graph

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

While all `geoms` takes mapping functions, they cant always contain the
same type of `asthetics`. For example, a line graph can not contain
shapes or different colors. However, for line graphs you can for example
set the `linetype` like this:

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

Here the function separates the car dependent on the drivetrain type
(`drv`). We can also do this, and add points to the graph to make it
clearer, containing two `geoms` in the same graph, like this.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

The problem here is however that you essentially write the code two
times. An alternative is to write code like this:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y= hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(mapping = aes(linetype = drv))
```

Here you can also see that it is possible to place mapping functions
inside only one geom and filter data, showing only some of it.

### 5.2 - Exercises

#### **Question 1 - What geom would you use to draw a line chart, boxplot, hisogram, or area chart?** 

**Answer;**

The following:

-   Linechart = `geom_smooth()`
-   Boxplot = `geom_boxplot()`
-   Histogram = `geom_histogram()`
-   Area chart = `geom_area()`

#### **Question 2 - Run this code in your head. And predict outcome**

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

**Answer;** I predict that the below code with make a colored scatter
plot of `displ` and `hwy` that is colored according to what type of
drivesystem the car has (`drv`). Furthermore, there will be a line graph
(w.o. SE) layered on top of this.

#### **Question 3 - What does show.legend = FALSE do? What happens if you remove it?**

**Answer:** It removes the legends box that shows what the different
`aesthetics` do.

#### **Question 4 - What does the `SE` argument in `geom_smooth()` do?**

**Answer:** It decides whether there should be a standard error line.

#### **Question 5 - Will these two graphs look different? Why, why not?**

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

vs

```{r}
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

**Answer:** They will going to look exactly the same. The first graph
defines mappings and data combined while the second does it
individually. Nevertheless, the variables are the identical, hence, the
graphs will be identical.

#### **Question 6 - Recreate some R code that generates some plots I see in the book**

**Answer:**

[Plot 1.]{.ul}

Here I need a scatter plot of `hwy` and `displ` with a line graph
without SE lines (blue).

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth(se = FALSE, color = "blue")
```

[Plot 2:]{.ul}

Here I need a scatter plot of `hwy` and `displ` with a line graph with
group mapping on `drv`. Still without SE lines (blue). Also, as the
legend will appear with grouping, I need to set `show.legend = FALSE`.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth(se = FALSE, color = "blue", aes(group = drv), show.legend = FALSE)
```

[Plot 3:]{.ul}

Here I need a scatter plot of `hwy` and `displ` with aesthetic mapping
of coloring on `drv`. Furthermore, I need a line graph with group
mapping on `drv` with color. Still without SE lines (blue).

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() + 
  geom_smooth(aes(group = drv), se = FALSE)
```

[Plot 4:]{.ul}

Here I need a scatter plot of `hwy` and `displ` with aesthetic mapping
of coloring on `drv`. Furthermore, I need a line graph with the color
blue. Still without SE lines (blue).

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) + 
  geom_smooth(se = FALSE)
```

[Plot 5:]{.ul}

Here I need a scatter plot of `hwy` and `displ` with aesthetic mapping
of coloring on `drv`. Furthermore, I need a line graph with group
mapping on `drv` with `linetype`. Still without SE lines (blue).

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) + 
  geom_smooth(se = FALSE, color = "blue", aes(linetype = drv))
```

[Plot 6:]{.ul}

Here I need a scatter plot of `hwy` and `displ` with aesthetic mapping
of coloring on `drv`.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() 
 
```

## 6 - Statistical Transformation

### 6.1 - Notes

Some plots automatically calculate new values for your plots. An example
is the `diamonds` data set where there is no count variable. When making
a bar plot a count is calculated:

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

When doing this, the plots bins the data, and counts the number/count
that falls within each bin. This occurs for bar charts, histograms, and
frequency polygons.

Other automatically calculated stuff can include smoothness model, and
summary from boxplots.

The algorithms used for these transformations are called a `stat` (short
for statistical transformation).

You can see/view which `stat` that `geoms` uses by using `?`.

```{r}
?geom_bar
```

Here it says:

> geom_bar() uses stat_count() by default.

You can sometimes use `stats` interchangeability with `geoms`. The
previous graph can also be made like this.

```{r}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

There are three reasons you might choose to use `stats` over `geoms`.

1.  Override default stat (i.e. change from count to identity)

```{r}
demo <- tribble (
  ~a,       ~b,
  "bar_1",  20,
  "bar_2",  30,
  "bar_3",  40,
)

ggplot(data = demo) +
  geom_bar(mapping = aes(x = a, y = b), stat = "identity")
```

2.  Override default mapping. You might want to show something other
    than a normal Y column, for example proportion.

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

3.  You can add more summary data to the plot.

```{r}
ggplot(data = diamonds) +
  stat_summary(mapping = aes(x = cut, y = depth), fun.min = min, fun.max = max, fun = median)
```

### 6.2 - Exercises

#### **Question 1 - What is the default geom associated with stat_summary()? How could you rewrite the previous plat using geom instead of stat?**

I will first look at the documentation for stat_summary()

```{r}
?stat_summary
```

Here it says we can use geom_pointrange. Lets try that.

```{r}
ggplot(data = diamonds) +
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary",
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median)

```

**2 - What does geom_col() do? How is it different to geom_bar()?**

Lets find out.

```{r}
?geom_col()
```

Geom_bar makes the height of the bar proportional to number of cases in
each group. If you want them to be reperesent values in the acctual
data, geom_col() does this. I.e. geom_bar() uses stat_count, while
geom_col() uses stat_identity.

**3 - Most geoms and stats come in pairs that are almost always used in
concert. What does these pairs have in common?**

We should read documentation.

```{r}
?stat_bin()
```

They often have similar names. They often do the same thing. Also, they
use each other for the same thing. I.e. geom uses a stat, and a stat
shows it using a geom. Dont really know why there are two types. Maybe
to make it easier in edgecases.

**4 - What variables does stat_smooth() compute? What parameter controls
its behavior?**

```{r}
?stat_smooth()
```

Stat smooth compute the SE using the predictdf function (usually). Also,
y, ymin, and yma. The parameters that controls its behavior is method,
formula, and na.rm.

**5 - What is the problems with these graphs?**

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop.., group = 1))
```

The problem here is that they are all one. We need to set group to one.
Prop is a groupwise proportion, and as such we need to have a group.

# 7 - Position Adjustments

## 7.1 - Notes

You can fill bar charts using the color aesthetic.

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, color = cut))
```

Or better yet

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))
```

You can also fill stuff in different variables like this:

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

You can use position function to set how the different colors should be
positioned. In the below example they are positioned based on their
identity.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "identity")
```

Alpha stand for ocupizaty.

You can also set a fill position

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill")
```

Here we fill the entire Y axis. This makes it easier to compare
propoprtions of clarity inside the same sample.

Lastly we can use the dodge function to make objects overlap.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "dodge")
```

There are also more types of adjustements that are more useful for other
types of geoms. For example, if points overlap (called overplotting) you
can avoid this by using the jitter function. This function adds some
random noise to each variable so that you are able to see them all.

This would in theory make your graph less accurate at small scales,
however, more accurate at large scales.

YOu can also use geom_jitter() to do this quickly.

Like this:

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

## 7.1 - Exercises

**1 - What is the problem with this plot and how can you fix it?**

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point()
```

The problem is that stuff overlap, as there are only that many ctys and
hwy combinations. This can be fixed by adding jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter()
```

**2 - What parameters to geom_citter() control the amount of
jittering?**

We will look at the documentations.

```{r}
?geom_jitter
```

Height and width is added to jitter. By default this is 40% of the
resolutinon of the data.

**3 - Compre and contrast geom_jitter with geom_count**

Lets make a geom count graph

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```

Geom count adds the count to the graph, increasing the size (or other
variables) dependent of how many items are in that exact positon.

**4 - What is the default position adjustment for geom_boxplot? Create a
visulization of mpg do demonstrate this**

Lets first look at the documentation.

```{r}
?geom_boxplot
```

The default position adjustement is "dodge2", which is.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_boxplot(position = "dodge2")
```

# 8 - Coordinate systems

## 8.1 - Notes

Coordinate systems are the most complicated part of ggplot2. The default
coordinate system is something called cartesian coordinate. This is
where x and y is independent. There is also a bunh of other stuff that
is usefull

-   Coord_flip = Switches x and y axis.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
```

-   coord quickmap = Sets aspect ratio correctly for maps (spartial
    data).

```{r}
install.packages("maps")
nz <- map_data("nz")

ggplot(data = nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
  coord_map()
  
```

-   Coords_polar = makes polar coordinate.

```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut), show.legend = FALSE, width = 1) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

## 8.2 - Exercises

**1 - Turn a stacked bar chart into pie chart using coord_polar**

```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut), show.legend = FALSE, width = 1) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

**2 - What does labs() do?**

Read the documentation.

```{r}
?labs
```

Labs modifies axis legend and plot labels.

**3 - What's the difference between coord_quickmap() and coord_map()?**

```{r}
?coord_quickmap
?coord_map
```

Quickmap is a approximation that does preserve straight lines.

```{r}
install.packages("maps")
nz <- map_data("nz")

ggplot(data = nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_map()

```

**4 - What does the following plot tell you? Why is coord_fixed
important? And what does geom_abline() do?**

Plot:

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed()
```

This plot tells us that there is a linear relationship between cty and
hwy.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() 
```

Without cord fixed, the x and y axis is no longer equal, making it
somehwat more difficult to see. I.e. one unit on the x axis is one unit
on the y axis when using coord_fixed().

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  coord_fixed() 
```

Geom_abline creates a refference line line.

# 9 - The layered grammer of graphics

## 9.1 - Notes

ggplot template:

```{r}
ggplot(date = <DATA>) +
  <GEOM>(mapping = aes(<MAPPINGS>), stat = <STAT>, position = <POSITION>) +
  <CORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

This template can build all types of graphs!
