---
title: "Chapter 3 - Data Visulisation"
output: html_notebook
---
# Load packages
```{r}
library(tidyverse)
```


# 3.1 - Introduction
## 3.1.2 - Notes
Only talks about prerequisites

# 3.2 - First steps
## 3.2.1 - Notes
We are using the MPG data set. This comes with the tidyverse package.

A standard graphing template is as follows.

```{r}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## 3.2.2 - Excersises
**1 - Run ggplot on mpg. What do you see?**

I run ggplot on mpg:

```{r}
ggplot (data = mpg)
```

Doing this I see nothing. I need to define what I want to see

**2 - How many rows and columns are in mpg?**

I can find this using the str function, or altenativly glimpse. Or just writing the data in the console.

```{r}
str(mpg)
glimpse(mpg)
mpg
```


It is 234 rows and 11 columns.

**3 - what does the dr variable describe?**

I can find this by looking trough the documentation.
```{r}
?mpg
```

drv describes whether the car is f (front wheel drive), r (rear wheel drive), or 4 (4wd)

**4 - make a scatterplot of hwy vs cyl**

I can do that using the following.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```

**5 - What happens if you make a scatterplot of class vs drv? Why is it not useful?**

This happens:
```{r}
ggplot(data = mpg) +
  geom_point(mapping = (aes(x = class, y = drv)))
```

This is not useful as the type of wheel setup (drv) is not dependent on type of car. One would expect.


# 3.3 - Aesthetic mappings
## 3.3.1 - Notes
Lets you highlight specific stuff. Astetics includes things like size, shape and color. One can use aestetic mapping like this.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Doing this will automatically assign a unique "level" of asthetic (color) tp each unique value in the chosen variable (in this case: car class).

Here is the same example, using size instead.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
 This gives us a different asthetic, however, in this case it looks real messy. Other things we can map to is:
 1. alpha = transperasy
 2. shape = shape
 
 We can also set astetics manually like this.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
Notice that the color is outside of the x and y parentheses, but still inside the aes.
 
You can look up information on the types of shapes that are avaliable in R. There is a total of 25.
 
## 3.3.2 - Excersises
**1 - What's gone wrong with this code? Why are the points not blue?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```
Answer: Because the decleration for blue points are inside the x and Y parentheses, which creates a new variable in MPG called blue, which R attempts to color.

**2 - Which variables in mpg are chatogorical and continious?**

We can figure this out by looking at the documentation.
```{r}
?mpg
```

It is somewhat difficult to look at the documentation. It is best to look at a printed data frame.
```{r}
mpg
```

Here we can see <chr> which stands for chategorical and <dbl> or <int> which stand continuous. dbl is double precision vectors, or double - from C. I.E. a long float.

**3 - Map a continuous variable to color, size, and shape and look how they behave compared to chategorical values**

I 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = year, size = cyl))
```
First it tells me that a continuous variable cant be mapped to shape. This makes sense.

As for colour, it creates a gradient when used with continous variables. Size also work with continous variables, however, is not as good and rather scales the variables instead of having discrete sizes.


**4 - What happens if you map the same variable to multiple asthetics?**

Lets try.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = trans, shape = trans))
```
Both things are mapped.

**5 - What does the stroke aesthetic do? What shapes does it work with?**

I can see this by looking at the documentation of geom_point.
```{r}
?geom_point
```


Stroke can modify the with of the border.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = year, size = cyl, stroke = cyl))
```
It can also do this:
```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)
```


**6 - What happens if you map an aesthetic to something other than a variable name?**

Lets find out

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = hwy, color = displ < 5))
```

You can get logical statements when doing this.

# 3.4 - Common problems
## 3.4.1 - Notes

## 3.4.1 - Excerises

